---
title: "The fastest way to compile table 1?"
author: "Max Gordon"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    css: custom.css
vignette: >
  %\VignetteIndexEntry{Gmisc for generating descriptive statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

The purpose of the first table in a medical paper is most often to describe your population. In an RCT the table frequently compares the baseline characteristics between the randomized groups, while an observational study will often compare exposed with unexposed. In this vignette I will show how I use the functions to quickly generate a descriptive table.

We will use the mtcars dataset and compare the groups with automatic transmission to those without. First we will start by loading the dataset and labeling.


```{r, message=FALSE}
data(mtcars)
# For labelling we use the label()
# function from the Hmisc package
library(Hmisc)

label(mtcars$mpg) <- "Gas"
units(mtcars$mpg) <- "Miles/(US) gallon"

label(mtcars$wt) <- "Weight"
units(mtcars$wt) <- "10<sup>3</sup> kg" # not sure the unit is correct

mtcars$am <- factor(mtcars$am, levels=0:1, labels=c("Automatic", "Manual"))
label(mtcars$am) <- "Transmission"

mtcars$gear <- factor(mtcars$gear)
label(mtcars$gear) <- "Gears"

# Make up some data for making it slightly more interesting
mtcars$col <- factor(sample(c("red", "black", "silver"),
                     size=NROW(mtcars), replace=TRUE))
label(mtcars$col) <- "Car color"
```

## The basics of getDescriptionStatsBy

The function **getDescriptionStatsBy** is a simple way to do basic descriptive statistics. Mandatory inputs are the first two, *x* and *by*:

```{r}
library(Gmisc)
getDescriptionStatsBy(x = mtcars$mpg, 
                      by = mtcars$am)
```

If we prefer median we can simply specify the statistic used with continuous variables:

```{r}
getDescriptionStatsBy(x = mtcars$mpg, 
                      by = mtcars$am,
                      continuous_fn = describeMedian)
```

As I always have the same *by*-variable and have several customization, I often wire a small wrapper:

```{r}
getTable1Stats <- function(x, digits = 0, ...){
  getDescriptionStatsBy(x = x, 
                        by = mtcars$am,
                        digits = digits,
                        continuous_fn = describeMedian,
                        header_count = TRUE,
                        ...)
  
}
getTable1Stats(mtcars$mpg)
```

The dot-argument (...) is useful for adding additional customization:

```{r}
getTable1Stats(mtcars$mpg, use_units = TRUE)
```

## Combining with the **htmlTable**

Lists are convenient and allow a lot of customization, together with the **mergeDesc** it is easy to generate a table one:

```{r, results='asis'}
t1 <- list()
t1[["Gas"]] <-
  getTable1Stats(mtcars$mpg)
  
t1[["Weight&dagger;"]] <-
  getTable1Stats(mtcars$wt)

t1[["Color"]] <- 
  getTable1Stats(mtcars$col)

# If we want to use the labels set in the beginning
# we add an element without a name
t1 <- c(t1,
        list(getTable1Stats(mtcars$gear)))

htmlTable(mergeDesc(t1),
          css.rgroup = "",
          caption  = "Basic descriptive statistics from the mtcars dataset",
          tfoot = "&dagger; The weight is in 10<sup>3</sup> kg")
```

Event though p-values are discouraged in the table 1, it is [not uncommon](http://cpr.sagepub.com/content/19/2/231.full). I have added basic statistics consisting of Fisher's exact test for proportions and Wilcoxon rank sum test for continuous values. Note that the Wilcoxon only works with two groups.

```{r, results='asis', warning=FALSE}
# A little more advanced input
mtcars$mpg_w_missing <- mtcars$mpg
mtcars$mpg_w_missing[sample(1:NROW(mtcars), size=5)] <- NA
mtcars$wt_w_missing <- mtcars$wt
mtcars$wt_w_missing[sample(1:NROW(mtcars), size=8)] <- NA

t1 <- list()
t1[["Gas"]] <-
  getTable1Stats(mtcars$mpg_w_missing, statistics=TRUE)
  
t1[["Weight&dagger;"]] <-
  getTable1Stats(mtcars$wt, statistics=TRUE)

t1[["Color"]] <- 
  getTable1Stats(mtcars$col, statistics=TRUE)

# If we want to use the labels set in the beginning
# we add an element without a name
t1 <- c(t1,
        list(getTable1Stats(mtcars$gear, statistics=TRUE)))

htmlTable(mergeDesc(t1),
          css.rgroup = "",
          caption  = "Basic descriptive statistics from the mtcars dataset",
          tfoot = "&dagger; The weight is in 10<sup>3</sup> kg")
```
