---
title: "Table mania"
author: "Max Gordon"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Table mania}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

Introduction
============

Tables are an essential part of publishing, well... anything. I therefore want to explore the options available for generating these in markdown. It is important to remember that there are two ways principal ways of generating tables in markdown:

1. Markdown adapted table
2. HTML tables

Mardkown tables
===============

Raw tables
----------

A markdown table is fairly straight forward and are simple to manually create. Just write the plain text below:

<pre>
1st Header  | 2nd Header
----------- | -------------
Content A   | Content B
Content C   | Content D
</pre>

And you will end up with this beauty:

1st Header  | 2nd Header
----------- | -------------
Content A   | Content B
Content C   | Content D

The `knitr::kable` function
---------------------------

Now this is not the R way, we want to use a function that does this. The **knitr** comes with a table function well suited for this, **kable**:

```{r, results='asis'}
# First generate the corresponding
# matrix to the above example
output <- 
  matrix(sprintf("Content %s", LETTERS[1:4]),
         ncol=2, byrow=TRUE)
colnames(output) <- 
  c("1st header", "2nd header")
rownames(output) <- 
  c("1st row", "2nd row")

library(knitr)
kable(output, 
      caption="A test table", 
      align = c("c", "r"))
```

The advantage with the `kable` function is that it outputs true markdown tables and these can through the [pandoc](http://johnmacfarlane.net/pandoc/README.html#tables) system be converted to any document format. Some of the downsides are:

* Lack of adding row groups and column groups
* No control over cell formatting
* No control over borders
* ...

HTML tables
===========

Tables are possibly the most tested HTML-element out there. In early web design this was the only feature that browsers handled uniformly, and became therefore for a long period the standard way of doing layout. They are thereby an excellent template for generating advanced tables in statistics. There are currently three different implementations that I've encountered, the **xtable**, the **format.tables**, and my own **htmlTable** function. 

The `xtable`-package
--------------------

The `xtable` is a solution that delivers both html and LaTeX. The syntaxt is very similar to `kable`:

```{r, results='asis'}
library(xtable)
print(xtable(output, 
             caption="A test table", 
             align = c("l", "c", "r")), 
      type="html")
```

The downside with the function is that you need to change output depending on your target and there is not that much advantage compared to `kable`.

The `format.tables`-package
-------------------------

The Swedish Pension Agency has been a pioneer here in Sweden when it comes to implementing R in their daily work. They currently produce their annual report in markdown, a very impressive piece of work. As part of this Cedric Perriard has developed a table solution that allows both html and LaTeX output, `format.tables`. The package needs to be installed from GitGub:

```{r, eval=FALSE}
# If you haven't installed it before, you need devtools
install.packages("devtools")
devtools::install_github(repo = "format.tables",
                         username = "SwedishPensionsAgency")
```

It allows both simple and advanced formatting, here is a basic example:

```{r, results='asis'}
library(format.tables)

x <- format.tables(data=head(cars, 3), 
                    styles=rep("plain", 3), 
                    column.names = names(cars), 
                    names.style = "header", 
                    header=list(caption = "Caption", subcaption="subcaption")
                    )
cat(x$render(type="html"))
```

The `htmlTable` in the `Gmisc`-package
--------------------------------------

I developed the `htmlTable` inorder to get tables matching those available in top medical journals. After finding no html-alternative to the `Hmisc::latex` function on [Stack Overflow](http://stackoverflow.com/questions/11950703/html-with-multicolumn-table-in-markdown-using-knitr) I wrote a basic function allowing column spanners and row groups. Below is a basic example on these two:

```{r, results='asis'}
output <- 
  matrix(paste("Content", LETTERS[1:16]), 
         ncol=4, byrow = TRUE)

library(Gmisc)
htmlTable(output,
          rgroup = c("Group A-H",
                     "Group I-P"),
          n.rgroup = c(2,2),
          rnames = paste(c("1st", "2nd",
                           "3rd", "4th"), "row"),
          cgroup = c("Cgroup 1", "Cgroup 2&dagger;"),
          n.cgroup = c(2,2), 
          caption="Basic table with both column spanners (groups) and row groups",
          tfoot="&dagger; A table footer commment"
          )
```

More *raw* markdown tables
========================

There are a few more text alternatives available when desiging tables. I included these from the manual for completeness.

<pre>
| Right | Left | Default | Center |
|------:|:-----|---------|:------:|
|   12  |  12  |    12   |    12  |
|  123  |  123 |   123   |   123  |
|    1  |    1 |     1   |     1  |

: Demonstration of pipe table syntax.
</pre>

| Right | Left | Default | Center |
|------:|:-----|---------|:------:|
|   12  |  12  |    12   |    12  |
|  123  |  123 |   123   |   123  |
|    1  |    1 |     1   |     1  |

  : Demonstration of pipe table syntax.

<pre>
: Sample grid table.

+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| Bananas       | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - tasty            |
+---------------+---------------+--------------------+
</pre>

: Sample grid table.

+---------------+---------------+--------------------+
| Fruit         | Price         | Advantages         |
+===============+===============+====================+
| Bananas       | $1.34         | - built-in wrapper |
|               |               | - bright color     |
+---------------+---------------+--------------------+
| Oranges       | $2.10         | - cures scurvy     |
|               |               | - tasty            |
+---------------+---------------+--------------------+