---
title: "Introduction to transitionPlot"
author: "Max Gordon"
date: "Saturday, January 17, 2015"
output:
  html_document:
    toc: yes
---

Introduction
============

The transitionPlot function aims at illustrating the transition between classes. The original intent was to show how (self-administered the Charnley classification behaves)[http://informahealthcare.com/doi/full/10.3109/17453674.2014.931199] before and after surgery. The plot is a fancier version than what can be achiewed using packages such as (*diagram*)[http://informahealthcare.com.proxy.kib.ki.se/doi/full/10.3109/17453674.2014.931199] but at the cost of flexibility. The current function only allows to show a transition from one state to the next.

Basic use
=========

We will start by simulating some data. Each observation has a sex and a Charnley class (A, B, or C). The transition is then dependent on both the sex and the Charnley class.

```{r}
set.seed(1)
library(magrittr)
n <- 100
data <- 
  data.frame(
    Sex = sample(c("Male", "Female"),
                 size = n,
                 replace = TRUE),
    Charnley_class = sample(c("A", "B", "C"), 
                            size = n, 
                            replace = TRUE))
prob <- data.frame(
  A = 1/3 +
    (data$Sex == "Male") * .25 +
    (data$Sex != "Male") * -.25 +
    (data$Charnley_class %in% "B") * -.5 +
    (data$Charnley_class %in% "C") * -2 ,
  B = 1/3 +
    (data$Sex == "Male") * .1 + 
    (data$Sex != "Male") * -.05 +
    (data$Charnley_class == "C") * -.5,
  C = 1/3 +
    (data$Sex == "Male") * -.25 +
    (data$Sex != "Male") * .25)

# Remove negative probabilities
prob <- t(apply(prob, 1, function(x) {
  if (any(x < 0)){
    x <- x - min(x) + .05
    }
  x
  }))

data$Charnley_class_after <- 
  apply(prob, 1, function(p)
    sample(c("A", "B", "C"), 
           size = 1, 
           prob = p)) %>%
  factor(levels = c("A", "B", "C"))

# Create the transition matrix that 
# is the basis for the transition plot
trn_mtrx <-
  with(data,
       table(Charnley_class, 
             Charnley_class_after))
  
rm(prob)
```

The most simple use is with a straight table object:

```{r}
library(Gmisc)
transitionPlot(trn_mtrx)
```

The out of the box arrows are rather in my opinion not well suited and I have therefore developed a few alternatives:

```{r}
library(Gmisc)
data %>%
  with(table(Charnley_class, 
             Charnley_class_after)) %>%
  transitionPlot(type_of_arrow = "simple")
```

```{r}
library(Gmisc)
data %>%
  with(table(Charnley_class, 
             Charnley_class_after)) %>%
  transitionPlot(type_of_arrow = "gradient", main = "Gradient")
```

Using 3D matrices
=================

We can enhance the (data-ink ratio)[http://www.infovis-wiki.net/index.php/Data-Ink_Ratio] by adding a color split for the boxes. This works only for factors with 2 levels such as sex in our example.

```{r}
trn_mtrx_3D <-
  with(data,
       table(Charnley_class, 
             Charnley_class_after,
             Sex))

transitionPlot(trn_mtrx_3D, 
               fill_start_box = c("darkblue", "darkgreen"),
               type_of_arrow = "gradient", 
               main = "Gradient")
```

