---
title: "Introduction to the Transition-class"
author: "Max Gordon"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Gmisc for generating descriptive statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

Introduction
============

The *Transition*-class is the advanced version of transitionPlot that aims at illustrating the transition between classes over time. The original intent was to show how [self-administered the Charnley classification](http://informahealthcare.com/doi/full/10.3109/17453674.2014.931199) behaves before and after surgery. The plot is a fancier version than what can be achiewed using packages such as [*diagram*](http://informahealthcare.com.proxy.kib.ki.se/doi/full/10.3109/17453674.2014.931199) but at the cost of flexibility. The current function only allows to show transitions from one state to the next and there is generally no going back.

Generate some data
==================


We will start by simulating some data similar to my article. Each observation has a sex and a Charnley class (A, B, or C). The transition is then dependent on both the sex and the Charnley class.

```{r}
set.seed(1)
library(magrittr)
n <- 100
data <- 
  data.frame(
    Sex = sample(c("Male", "Female"),
                 size = n,
                 replace = TRUE,
                 prob = c(.4, .6)),
    Charnley_class = sample(c("A", "B", "C"), 
                            size = n, 
                            replace = TRUE))
getProbs <- function(Chrnl_name){
  prob <- data.frame(
    A = 1/3 +
      (data$Sex == "Male") * .25 +
      (data$Sex != "Male") * -.25 +
      (data[[Chrnl_name]] %in% "B") * -.5 +
      (data[[Chrnl_name]] %in% "C") * -2 ,
    B = 1/3 +
      (data$Sex == "Male") * .1 + 
      (data$Sex != "Male") * -.05 +
      (data[[Chrnl_name]] == "C") * -2,
    C = 1/3 +
      (data$Sex == "Male") * -.25 +
      (data$Sex != "Male") * .25)
  
  # Remove negative probabilities
  t(apply(prob, 1, function(x) {
    if (any(x < 0)){
      x <- x - min(x) + .05
      }
    x
    }))
}
  
Ch_classes <- c("Charnley_class")
Ch_classes %<>% c(sprintf("%s_%dyr", Ch_classes, c(1,2,6)))
for (i in 1:length(Ch_classes)){
  if (i == 1)
    next;

  data[[Ch_classes[i]]] <- 
    apply(getProbs(Ch_classes[i-1]), 1, function(p)
      sample(c("A", "B", "C"), 
             size = 1, 
             prob = p)) %>%
    factor(levels = c("A", "B", "C"))
}
```

Basic use
=========
```{r, echo=FALSE, }
knitr::opts_chunk$set(dev.args=list(type="cairo"), 
                      message=FALSE, 
                      warnings=FALSE)
knitr::opts_chunk$set(fig.height = 5, fig.width=5)
```

The most simple use is to just supply the output from the `table()` call:

```{r}
library(Gmisc)
transitions <- getRefClass("Transition")$new(table(data$Charnley_class, data$Charnley_class_1yr), 
                                                  label=c("Before surgery", "1 year after"))
transitions$render()
```

Adding customizations are rather straight forward by setting the different field arguments:

```{r}
transitions <- getRefClass("Transition")$new(table(data$Charnley_class, data$Charnley_class_1yr), 
                                                  label=c("Before surgery", "1 year after"))
transitions$title <- "Charnley class in relation to THR"
transitions$arrow_type <- "simple"
transitions$box_label_pos <- "bottom"
transitions$render()
```

Adding a third dimension
------------------------

Underlying proportions can be visualized by splitting the box colors and blending the colors in the gradient arrows. The color blend is explained in the colorbar element at the bottom:

```{r}
transitions <- getRefClass("Transition")$new(table(data$Charnley_class, data$Charnley_class_1yr, data$Sex), 
                                                  label=c("Before surgery", "1 year after"))
transitions$title <- "Charnley class in relation to THR"
transitions$clr_bar <- "top"
transitions$render()
```

Multiple transitions
====================

The major advantage with the **Transition**-class is that it allows for transitions at multiple time points. We do this by calling the **addTransitions** function as shown below:

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.height = 5, fig.width=7)
```

```{r}
transitions <- getRefClass("Transition")$new(table(data$Charnley_class, data$Charnley_class_1yr, data$Sex), 
                                                  label=c("Before surgery", "1 year after"))
transitions$title <- "Charnley class in relation to THR"
transitions$addTransitions(table(data$Charnley_class_1yr, data$Charnley_class_2yr, data$Sex), label="2 years after")
library(grid)
transitions$max_lwd <- unit(.05, "npc")
transitions$render()
```